---
title: "Mini Data Analysis Milestone 2"
output: md_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. Within a genus of Vancouver trees, does the average diameter and height of a tree differ between species? Do these variables differ between genera?
2. Are the different genera of trees spread uniformly across Vancouver or do certain genera of trees cluster within certain parts of Vancouver, such as by neighbourhoods?
3. Can we detect a difference in the average age of trees in different neighbourhoods of Vancouver? Does this correspond to the age of the neighbourhood?
4. In Vancouver, how often are trees planted in the presence of a curb? Does the presence of a curb next to where a tree is growing affect its diameter and height?
<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
### Research Question 1

#### Summarizing

```{r}
four_trees <- vancouver_trees %>%
  filter(genus_name %in% c("ACER", "PRUNUS", "FRAXINUS", "CARPINUS")) # filtering for trees of these four genera

four_trees_summary <- four_trees %>%
  group_by(genus_name) %>%
  summarize(range = max(diameter) - min(diameter), # calculating range of the diameters
            mean = mean(diameter),
            median = median(diameter),
            variance = var(diameter)) 

print(four_trees_summary)

```

#### Graphing

```{r}

acer_prunus_distribution <- vancouver_trees %>%
  filter(genus_name %in% c("ACER", "PRUNUS", "FRAXINUS", "CARPINUS")) %>%
  filter(diameter < 60) %>%  # filter for trees with diameters less than 80
  group_by(diameter) %>%
  ggplot(aes(x = diameter, fill = genus_name)) +
  geom_histogram(bins = 50, alpha = 0.5) + # creates histogram and 50 different groups to define the histogram
  geom_freqpoly(aes(color = genus_name), alpha = 0.9) + # adds line graph for tree diameters
  labs(x = "Diameter (in)", y = "Number of trees",
       title = "Distribution of the Diameters of Acer, Prunus, and Fraxinus Trees") +
  scale_x_continuous(breaks = seq(0, 60, by = 10)) + # setting the scale of the x axis
  scale_y_continuous(breaks = seq(0, 18000, by = 2000)) # setting the scale of the y axis

print(acer_prunus_distribution)

```

### Research Question 2

#### Summarizing

```{r}

genus_neighbourhood <- vancouver_trees %>%
  group_by(genus_name, neighbourhood_name) %>% # groups data by tree genus and neighbourhood
  summarize(count = n(), # count of genus of tree in a neighbourhood
            proportion = n() / nrow(.)) # proportion of genus of tree in a neighbourhood

print(genus_neighbourhood)

```

#### Graphing

```{r}
tree_map <- vancouver_trees %>%
  filter(genus_name %in% c("ACER", "FRAXINUS", "PRUNUS")) %>%
  ggplot(aes(x = longitude, y = latitude, color = genus_name)) + # graphs trees based on their longitude and latitude
  geom_point(size = 0.08, alpha = 0.4) + # creates points for each tree
  labs(x = "Longitude", y = "Latitude",
       title = "Position of Acer, Fraxinus, and Prunus Trees by Neighbourhood")

print(tree_map)
```

### Research Question 3

#### Summarizing

```{r}
tree_ages <- vancouver_trees %>%
  mutate(age_years = as.numeric(difftime(as.Date("2023-10-25"), date_planted, units = "days") / 365.25)) %>% # calculates age of tree in years using the date planted and makes a new column for this
  select(genus_name, species_name, neighbourhood_name, date_planted, age_years) # removes unnecessary columns
print(tree_ages)

tree_ages_summary <- tree_ages %>%
  group_by(neighbourhood_name) %>%
  summarize(age_range = max(age_years, na.rm = TRUE) - min(age_years, na.rm = TRUE), # calculates range of tree ages
            mean_age = mean(age_years, na.rm = TRUE),
            median_age = median(age_years, na.rm = TRUE),
            variance_age = var(age_years, na.rm = TRUE))
print(tree_ages_summary)

```
#### Graphing

```{r}

tree_age_graph <- tree_ages_summary %>%
  ggplot(aes(x = neighbourhood_name, y = mean_age)) +
  geom_bar(stat="identity", fill = "lightsteelblue3") + # creates bar graph for mean tree age
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) + # puts x-axis labels at a 50° angle to be readable
  labs(x = "Neighbourhood", y = "Mean Age",
       title = "Mean Age of Trees by Neighbourhood") +
  geom_point(aes(y = median_age), shape = 18, colour = "slateblue2", size = 4, alpha = 0.8) # adds points showing median tree age
  

print(tree_age_graph)

```

### Research Question 4

#### Summarizing

```{r}

curb_trees <- vancouver_trees %>%
  group_by(curb, height_range_id) %>% # groups data by whether a tree is planted by a curb and its height range
  summarize(count = n(), # count of trees
            proportion = n() / nrow(.)) # proportion of trees

print(curb_trees)

```

#### Graphing

```{r}
curb_trees_graph <- vancouver_trees %>%
  filter(diameter < 60) %>% # filters for trees with diameters less than 60 to remove outliers
  ggplot(aes(x = curb, y = diameter)) +
  geom_boxplot(alpha = 0.6, fill = "lightblue") + # creates boxplot
  geom_violin(alpha = 0.5, linewidth = 0, fill = "slateblue") # creates violin plot to better visualize the data

print(curb_trees_graph)
```

<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->

#### Research Question 1

After completing the operations for research question 1, I feel as though I am a little bit closer to answering my research question. I have determined that trees of the genus Prunus are wider on average, compared to Acer, Fraxinus, and Carpus trees. I have also looked at the median diameters for trees of these genera, which can help address my research question. However, I have not conduced these analyses on other genera of tree. I have also not looked between species or at height differences between different trees. To refine my research question, I would look into only genera or only species of trees. 

#### Research Question 2

I feel as though I have also answered research question 2 well, particularly in the "graphing" section. I was able to observe the distribution of Acer, Prunus, and Fraxinus trees within Vancouver. I am also able to see which genera of trees cluster in which neighbourhoods by looking at count and proportion in the "summarizing" section. However, it is hard to see specific clusters of trees on the graph. I also was not able to map the areas of different neighbourhoods on the graph. To refine the question, I could filter for trees with specific characteristics.

#### Research Question 3

Good progress has been made on answering research question 3. The mean and median ages of trees were calculated for each neighbourhood, and these values were also visualized on a graph using two geom layers. While the mean tree age is similar for each neighbourhood, some neighbourhoods such as Hastings-Sunrise and South Cambie stand out as having older trees. Other neighbourhoods such as Downtown, Kitsilano, and Strathcona appear to have younger trees. Despite this, I would still need to correlate the mean ages of trees to mean ages of neighbourhoods. To refine this question, I could also look into what street the trees are planted on. 
#### Research Question 4

While my operations bring my slightly closer to answering this research question, much more would remain to be done. For example, the boxplot shows a slight trend for trees planted by a curb to have a smaller diameter. However, this boxplot looks at all Vancouver trees as a whole. Trees are not sorted by genera, meaning that some trees will be naturally wider than others due to their genus. It is possible that trees that will not grow to a large diameter are purposefully planted by curbs, instead of curbs interfering with tree growth. Therefore, I would have to control for this factor.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->

The data in vancouver_trees is tidy data. This is because all of the variables are in columns, all observations for each variable are placed in rows, and single values for variables and observations are within their respective cells. For example, the first 8 variables tree_id, civic_number, std_street, genus_name, species_name, cultivator_name, common_name, and assigned are all columns. Each observation, which is one tree, represents one row. 

<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

#### Untidying the Data

To untidy my data, I used pivot_wider to move diameter values into new columns that correspond to the genus_name of the tree. This data is untidy as it results in many NA values, and observations for diameters are spread out across many columns.

```{r}

untidy_trees <- vancouver_trees %>%
  pivot_wider(names_from = genus_name,
               values_from = diameter) # pivots data wider such that genus_name becomes columns and values from diameter correspond to genus name, making the data untidy

print(untidy_trees)

```

#### Tidying the Data

To tidy my data, I used pivot_longer to return the data to its original state. This moves genus_name back into its own column with the genus name as the values. Diameter observations are also moved into one column. I then rearranged the order of the columns to match the original dataset.

```{r}

tidy_trees <- untidy_trees %>%
  pivot_longer(c(ULMUS:ALBIZIA),
               names_to = "genus_name",
               values_to = "diameter",
               values_drop_na = TRUE) %>% # pivots data longer such that genus_name and diameter become values in separate columns, making the data tidy again
  select(c("tree_id":"std_street", "genus_name", "species_name":"height_range_id", "diameter", "curb":"latitude")) # puts columns in the same order as they were previously

print(tidy_trees)
  
```

<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1. Within a genus of Vancouver trees, does the average diameter and height of a tree differ between species? Do these variables differ between genera?

2. Are the different genera of trees spread uniformly across Vancouver or do certain genera of trees cluster within certain parts of Vancouver, such as by neighbourhoods?

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->

I chose research question 1 as I believe there is an interesting difference in the diameters of trees between genera. I would like to explore these differences more in Part 3.

I chose research question 2 as I am still very interested in how different genera of trees are planted in different areas of Vancouver.

<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

<!--------------------------- Start your work below --------------------------->

### Version of Data to Answer the 2 Questions

```{r}

tree_diameters_areas <- vancouver_trees %>%
  select(tree_id, genus_name, on_street_block, on_street, neighbourhood_name, height_range_id, diameter, longitude, latitude) %>% # dropping irrelevant columns
  filter(!is.na(longitude), !is.na(latitude)) %>% # removes longitude and latitude values that are NA
  mutate(area = pi * (diameter / 2)^2) %>% # creating a new column that is the area of the trees, calculated from its diameter
  arrange(desc(diameter)) # arranging data by descending diameter

print(tree_diameters_areas)

```

<!----------------------------------------------------------------------------->

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: Within a genus of Vancouver trees, does the average diameter and height of a tree differ between species? Do these variables differ between genera?

**Variable of interest**: Average diameter

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->

```{r}

anova_trees <- vancouver_trees %>%
  select(genus_name, diameter)

model <- lm(diameter ~ as.factor(genus_name), data = anova_trees) # regress diameter on a dummy variable for each genus type

anova_result <- anova(model) # conducting anova on the regression model
print(anova_result)

```

<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->

```{r}

fitted_data <- broom::tidy(model) # looks at a summary of the different components of the model
print(fitted_data)

```

<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->

```{r}

library(readr)
library(here)

write_csv(four_trees_summary, file = here::here("output", "tree_diameter_summary.csv")) # writes a summary table from four_tree_summary into a folder named "output"

```

<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->

```{r}

saveRDS(fitted_data, file = here("output", "fitted_data.rds")) # writes fitted_data into an R binary file

fitted_data_read <- readRDS(here("output", "fitted_data.rds")) # loads the R binary file fitted_data.rds
fitted_data_read

```

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.
